Co-moving stars in \textsl{Gaia DR1}

### 2016-10-03 paper/ms.tex
no parallax
change title and add authors

\title{Widely separated, comoving stars in the \textsl{Gaia DR1 TGAS} sample}
\author{APW, SO, DWH, DNS, DSTN, others}

% Affiliations
% \newcommand{\pu}{1}
% \newcommand{\adrn}{2}
% \newcommand{\ccpp}{3}
% \newcommand{\mpia}{4}
% \newcommand{\uw}{5}
% \newcommand{\sagan}{6}

% \altaffiltext{\pu}{Department of Astrophysical Sciences,
%                    Princeton University, Princeton, NJ 08544, USA}
% \altaffiltext{\adrn}{To whom correspondence should be addressed:
%                      \texttt{adrn@princeton.edu}}
% \altaffiltext{\ccpp}{Center for Cosmology and Particle Physics,
%                      Department of Physics,
%                      New York University, 4 Washington Place,
%                      New York, NY 10003, USA}
% \altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie,
%                      K\"onigstuhl 17, D-69117 Heidelberg, Germany}
% \altaffiltext{\uw}{Astronomy Department, University of Washington,
%                    Seattle, WA 98195, USA}
% \altaffiltext{\sagan}{Sagan Fellow}

@@

@@

In what follows, all vectors are column vectors, unless we have transposed them.
We indicate whether a symbol represents a scalar, vector, or tensor
(matrix) by context, not typesetting.
Properties of the squared exponential:
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] - \Delta)\,\dd x\right]
  &=& +\frac{1}{2}\ln ||2\pi\,A|| -\Delta
  \quad ,
\end{eqnarray}
where $x$ and $\nu$ are $D$-dimensional vectors, $A$ is a positive definite
matrix, $\Delta$ is a scalar, and the integral is implicitly over all
of $D$-dimensional $x$-space.
We will need to complete squares.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] + \Delta
  &=& \frac{1}{2}\,\transp{x}\,\inv{A}\,x + \transp{x}\,B\,b + C
  \quad ,
\end{eqnarray}
where $B\,b$ is a $d$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \nu &=& A\,B\,b
  \\
  \Delta & = & C - \frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \quad .
\end{eqnarray}

Now in this case, we want to write down the probability density for the data
given the distances and velocity (the likelihood), and the prior pdf for
the velocity, and their product.
The prior pdf for the velocity is Gaussian:
\begin{eqnarray}
  p(v) &=& \normal(v\given 0,V)
  \\
  \ln p(v) &=& -\frac{1}{2}\,\ln||2\pi\,V|| -\frac{1}{2}\,\transp{v}\,\inv{V}\,v
  \\
  V &=& \sigma_v^2\,I
  \quad ,
\end{eqnarray}
where $v$ is the three-vector velocity, and $V$ is an isotropic
variance tensor.
Similarly, the likelihood for the data is also a Gaussian.
Before we write that down, we will make some definitions.
We construct a velocity-space data vector $y$ as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [(d_1\,\mu_{\alpha1}), (d_1\,\mu_{\delta1}),
                       (d_2\,\mu_{\alpha2}), (d_2\,\mu_{\delta2})] \quad \mbox{or}
  \\
  \transp{y} &\equiv& [(d_1\,\mu_{\alpha1}), (d_1\,\mu_{\delta1}), (v_{r1}),
                       (d_2\,\mu_{\alpha2}), (d_2\,\mu_{\delta2}), (v_{r2})]
  \quad ,
\end{eqnarray}
where we have given two versions, a 4-component and an 6-component
version (depending on whether there are measured radial velocities),
the subscripts 1, 2 refer to the first and second stars (arbitrarily
ordered) of the pair, we have multiplied the strict observables by
distances $d_1, d_2$, which we are permitted to do (because we are
\emph{conditioning} on $d_1, d_2$), and note the wacky minus-ones.
The vector $y$ is a column vector that we have transposed to make it
easy to write down.
Fundamentally, our H1 model (same velocity) is
\begin{eqnarray}
  y &=& M\,v + \mathrm{noise}
  \\
  \transp{M} &=& [\hat{\alpha_1}, \hat{\delta_1},
                  \hat{\alpha_2}, \hat{\delta_2}] \quad \mbox{or}
  \\
  \transp{M} &=& [\hat{\alpha}_1, \hat{\delta}_2, \hat{r}_1,
                  \hat{\alpha}_2, \hat{\delta}_2, \hat{r}_2]
  \quad,
\end{eqnarray}
where $M$ is a $6\times 3$ or $8\times 3$ design matrix, with zeros
in two rows, and unit vectors along the other rows.
The noise (in the $y$ vector) is drawn from a Gaussian with ($6\times 6$ or $8\times 8$)
covariance matrix $\Sigma$, constructed from the distances $d_1, d_2$ and
the \gaia-provided individual-object covariance matrices.
Given all these definitions, the likelihood function is
\begin{eqnarray}
  p(\data\given v,d_1,d_2) &=& d_1^3\,d_2^3\,\normal(y\given M\,v, \Sigma)
  \\
  \ln p(\data\given v,d_1,d_2) &=& 3\,\ln d_1 + 3\,\ln d_2
  -\frac{1}{2}\,\ln||2\pi\,\Sigma|| \nonumber \\ && \quad
  -\frac{1}{2}\,\transp{[y-M\,v]}\,\inv{\Sigma}\,[y-M\,v]
  \quad ,
\end{eqnarray}
where the factor of $d_1^3\,d_2^3$ converts to units of per \gaia\ data
from units of per $y$.

Now we multiply together (add in the log) and complete the square.
We get, in the original parameterization:
\begin{eqnarray}
  \inv{A} &=& [\transp{M}\,\inv{\Sigma}\,M+\inv{V}]
  \\
  \nu &=& A\,\transp{M}\,\inv{\Sigma}\,y
  \\
  \Delta &=& -3\,\ln d_1 -3\,\ln d_2
  +\frac{1}{2}\,\ln||2\pi\,\Sigma|| +\frac{1}{2}\,\ln||2\pi\,V|| \nonumber \\ && \quad
  +\frac{1}{2}\,\transp{y}\,\inv{\Sigma}\,y -\frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \quad ,
\end{eqnarray}
where there are sign issues that Hogg doesn't like to think about.
Finally, we marginalize out the velocity to get the marginalized
likelihood conditioned on the two distances $d_1, d_2$:
\begin{eqnarray}
  \ln p(\data\given d_1,d_2)
  &=& \frac{1}{2}\ln ||2\pi\,A|| -\Delta
  \quad .
\end{eqnarray}

The story for the H2 model (independent velocities) is very
similar. In this case, the marginalized likelihood is a product of
two independent integrals $Q$ each constructed as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [(d\,\mu_{\alpha}), (d\,\mu_{\delta})] \quad \mbox{or}
  \\
  \transp{y} &\equiv& [(d\,\mu_{\alpha}), (d\,\mu_{\delta}), (v_{r})]
  \\
  \transp{M} &=& [\hat{\alpha}, \hat{\delta}] \quad \mbox{or}
  \\
  \transp{M} &=& [\hat{\alpha}, \hat{\delta}_2, \hat{r}]
  \\
  \inv{A} &=& [\transp{M}\,\inv{\Sigma}\,M+\inv{V}]
  \\
  \nu &=& A\,\transp{M}\,\inv{\Sigma}\,y
  \\
  \Delta &=& -3\,\ln d
  +\frac{1}{2}\,\ln||2\pi\,\Sigma|| +\frac{1}{2}\,\ln||2\pi\,V|| \nonumber \\ && \quad
  +\frac{1}{2}\,\transp{y}\,\inv{\Sigma}\,y -\frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \\
  Q &=& \frac{1}{2}\ln ||2\pi\,A|| -\Delta
  \quad ,
\end{eqnarray}
where again there are two options for the dimensionality of $y$ and
$M$, and once the vectors and $\Sigma$ are properly truncated to one
star, everything else follows.

>>>>

\title{Co-moving stars in \textsl{Gaia DR1}}
\author{Adrian M. Price-Whelan\altaffilmark{\pu,\adrn},
        Semyeong Oh\altaffilmark{\pu},
        Timothy D. Morton\altaffilmark{\pu},
        David W. Hogg\altaffilmark{\ccpp,\mpia},
        David N. Spergel\altaffilmark{\pu,\cca}
}

% Affiliations
\newcommand{\pu}{1}
\newcommand{\adrn}{2}
\newcommand{\ccpp}{3}
\newcommand{\mpia}{4}
\newcommand{\cca}{5}

\altaffiltext{\pu}{Department of Astrophysical Sciences,
                   Princeton University, Princeton, NJ 08544, USA}
\altaffiltext{\adrn}{To whom correspondence should be addressed:
                     \texttt{adrn@princeton.edu}}
\altaffiltext{\ccpp}{Center for Cosmology and Particle Physics,
                     Department of Physics,
                     New York University, 4 Washington Place,
                     New York, NY 10003, USA}
\altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie,
                     K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{\cca}{Simons Center for Computational Astrophysics, ...,
                    New York, NY XXXXX, USA}

@@

\section{Data} \label{sec:data}

@@

In what follows, all vectors are column vectors, unless we have transposed them.
We indicate whether a symbol represents a scalar, vector, or tensor
(matrix) by context, not typesetting.
Properties of the squared exponential:
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] - \Delta)\,\dd x\right]
  &=& +\frac{1}{2}\ln ||2\pi\,A|| -\Delta
  \quad ,
\end{eqnarray}
where $x$ and $\nu$ are $D$-dimensional vectors, $A$ is a positive definite
matrix, $\Delta$ is a scalar, and the integral is implicitly over all
of $D$-dimensional $x$-space.
We will need to complete squares.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] + \Delta
  &=& \frac{1}{2}\,\transp{x}\,\inv{A}\,x + \transp{x}\,B\,b + C
  \quad ,
\end{eqnarray}
where $B\,b$ is a $d$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \nu &=& -A\,B\,b
  \\
  \Delta & = & C - \frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \quad .
\end{eqnarray}

Now in this case, we want to write down the probability density for the data
given the distances and velocity (the likelihood), and the prior pdf for
the velocity, and their product.
The prior pdf for the velocity is Gaussian:
\begin{eqnarray}
  p(v) &=& \normal(v\given 0,V)
  \\
  \ln p(v) &=& -\frac{1}{2}\,\ln||2\pi\,V|| -\frac{1}{2}\,\transp{v}\,\inv{V}\,v
  \\
  V &=& \sigma_v^2\,I
  \quad ,
\end{eqnarray}
where $v$ is the three-vector velocity, and $V$ is an isotropic
variance tensor.
Similarly, the likelihood for the data is also a Gaussian.
Before we write that down, we will make some definitions.
We construct a velocity-space data vector $y$ as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [(d_1\,\mu_{\alpha1}), (d_1\,\mu_{\delta1}),
                       (d_2\,\mu_{\alpha2}), (d_2\,\mu_{\delta2})] \quad \mbox{or}
  \\
  \transp{y} &\equiv& [(d_1\,\mu_{\alpha1}), (d_1\,\mu_{\delta1}), (v_{r1}),
                       (d_2\,\mu_{\alpha2}), (d_2\,\mu_{\delta2}), (v_{r2})]
  \quad ,
\end{eqnarray}
where we have given two versions, a 4-component and an 6-component
version (depending on whether there are measured radial velocities),
the subscripts 1, 2 refer to the first and second stars (arbitrarily
ordered) of the pair, we have multiplied the strict observables by
distances $d_1, d_2$, which we are permitted to do (because we are
\emph{conditioning} on $d_1, d_2$), and note the wacky minus-ones.
The vector $y$ is a column vector that we have transposed to make it
easy to write down.
Fundamentally, our H1 model (same velocity) is
\begin{eqnarray}
  y &=& M\,v + \mathrm{noise}
  \\
  \transp{M} &=& [\hat{\alpha_1}, \hat{\delta_1},
                  \hat{\alpha_2}, \hat{\delta_2}] \quad \mbox{or}
  \\
  \transp{M} &=& [\hat{\alpha}_1, \hat{\delta}_2, \hat{r}_1,
                  \hat{\alpha}_2, \hat{\delta}_2, \hat{r}_2]
  \quad,
\end{eqnarray}
where $M$ is a $6\times 3$ or $8\times 3$ design matrix, with zeros
in two rows, and unit vectors along the other rows.
The noise (in the $y$ vector) is drawn from a Gaussian with ($6\times 6$ or $8\times 8$)
covariance matrix $\Sigma$, constructed from the distances $d_1, d_2$ and
the \gaia-provided individual-object covariance matrices.
Given all these definitions, the likelihood function is
\begin{eqnarray}
  p(\data\given v,d_1,d_2) &=& d_1^2\,d_2^2\,\normal(y\given M\,v, \Sigma)
  \\
  \ln p(\data\given v,d_1,d_2) &=& 3\,\ln d_1 + 3\,\ln d_2
  -\frac{1}{2}\,\ln||2\pi\,\Sigma|| \nonumber \\ && \quad
  -\frac{1}{2}\,\transp{[y-M\,v]}\,\inv{\Sigma}\,[y-M\,v]
  \quad ,
\end{eqnarray}
where the factor of $d_1^2\,d_2^2$ converts to units of per \gaia\ data
from units of per $y$.

Now we multiply together (add in the log) and complete the square.
We get, in the original parameterization:
\begin{eqnarray}
  \inv{A} &=& [\transp{M}\,\inv{\Sigma}\,M+\inv{V}]
  \\
  \nu &=& -A\,\transp{M}\,\inv{\Sigma}\,y
  \\
  \Delta &=& -2\,\ln d_1 -2\,\ln d_2
  +\frac{1}{2}\,\ln||2\pi\,\Sigma|| +\frac{1}{2}\,\ln||2\pi\,V|| \nonumber \\ && \quad
  +\frac{1}{2}\,\transp{y}\,\inv{\Sigma}\,y -\frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \quad ,
\end{eqnarray}
where there are sign issues that Hogg doesn't like to think about.
Finally, we marginalize out the velocity to get the marginalized
likelihood conditioned on the two distances $d_1, d_2$:
\begin{eqnarray}
  \ln p(\data\given d_1,d_2)
  &=& \frac{1}{2}\ln ||2\pi\,A|| -\Delta
  \quad .
\end{eqnarray}

The story for the H2 model (independent velocities) is very
similar. In this case, the marginalized likelihood is a product of
two independent integrals $Q$ each constructed as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [(d\,\mu_{\alpha}), (d\,\mu_{\delta})] \quad \mbox{or}
  \\
  \transp{y} &\equiv& [(d\,\mu_{\alpha}), (d\,\mu_{\delta}), (v_{r})]
  \\
  \transp{M} &=& [\hat{\alpha}, \hat{\delta}] \quad \mbox{or}
  \\
  \transp{M} &=& [\hat{\alpha}, \hat{\delta}_2, \hat{r}]
  \\
  \inv{A} &=& [\transp{M}\,\inv{\Sigma}\,M+\inv{V}]
  \\
  \nu &=& A\,\transp{M}\,\inv{\Sigma}\,y
  \\
  \Delta &=& -2\,\ln d
  +\frac{1}{2}\,\ln||2\pi\,\Sigma|| +\frac{1}{2}\,\ln||2\pi\,V|| \nonumber \\ && \quad
  +\frac{1}{2}\,\transp{y}\,\inv{\Sigma}\,y -\frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \\
  Q &=& \frac{1}{2}\ln ||2\pi\,A|| -\Delta
  \quad ,
\end{eqnarray}
where again there are two options for the dimensionality of $y$ and
$M$, and once the vectors and $\Sigma$ are properly truncated to one
star, everything else follows.
